---
title: "New Insights on Poverty"
date: "January 29, 2016"
output: html_document
---

# Introduction

This lab is based Hans Rosling talks [New Insights on Poverty](https://www.ted.com/talks/hans_rosling_reveals_new_insights_on_poverty?language=en) and [The Best Stats You've Ever Seen](https://www.ted.com/talks/hans_rosling_shows_the_best_stats_you_ve_ever_seen). 

The lab uses data to answer specific question about global health and economics. The data contradicts commonly held preconceived nations. Let's start by testing your knowledge. For each of the six pairs of countries below, which country do you think  had the highest child mortality in 2015? Which pairs do you think are similar?

1. Sri Lanka or Turkey
2. Poland or South Korea
3. Malaysia or Russia
4. Pakistan or Vietnam
5. Thailand or South Africa

Now let's use data to answer the questions. There are a number of datasets available on the [website](in). You can visit the website and browse the datasets to find this one.

```{r load libraries,message=FALSE,echo=TRUE}
library(dplyr)
library(readr)
child_mortality <-read_csv("http://spreadsheets.google.com/pub?key=0ArfEDsV3bBwCcGhBd2NOQVZ1eWowNVpSNjl1c3lRSWc&output=csv")
```


By printing `dat` 

```{r,echo=FALSE}
child_mortality
```
we see that years are represented by columns and we can get these values for our countries using the following code:
```{r}
countries <- c("Sri Lanka","Turkey", "Poland", "South Korea", "Malaysia", "Russia","Pakistan","Vietnam","Thailand","South Africa")

rename(child_mortality, country=`Under five mortality`) %>% 
slice(match(countries,country)) %>% select(country,`2015`) 
```

From here we see that Turkey has a higher rate than Turkey, Poland higher than South Korea, Russia higher than Malaysia, Pakistan much higher than Vietnam and South Africa much higher than Thailand. 

Most people get these wrong. Why is this? In part it is due to our preconceived notion that the world is divided into two groups: Western world versus the third world characterized by long life in a small family
and short life in a large family respectively. Does the data support this view of two groups?

The necessary data to answer this question is also available from gapminder:

```{r}
life_expectancy  <- read_csv("http://spreadsheets.google.com/pub?key=phAwcNAVuyj2tPLxKvvnNPA&output=csv")
total_fertility  <- read_csv("http://spreadsheets.google.com/pub?key=phAwcNAVuyj0TAlJeCEzcGQ&output=csv")
```

We want to plot one variable versus the other. However we need to assure that we match the data correctly. To do this we will use the `tidyr` package. 

# Tidying data

Before we get started, note that the first column of all three datasets are incorrectly labeled. We fix this right away:

```{r}
names(child_mortality)[1] <- "country"
names(life_expectancy)[1] <- "country"
names(total_fertility)[1] <- "country"
```

Now we want each country/year combination to be a unit. We achieve this by using the `gather` function:


```{r}
library(tidyr)
child_mortality <- child_mortality %>% 
  gather(key = year, value = child_mortality,
         2:ncol(child_mortality), -country, convert=TRUE)
life_expectancy <- life_expectancy %>% 
  gather(key = year, value = life_expectancy,
         2:ncol(life_expectancy), -country, convert=TRUE)
total_fertility <- total_fertility %>% 
  gather(key = year, value = total_fertility,
         2:ncol(total_fertility), -country, convert=TRUE)
```

Now we can join any two columns 
```{r}
dat <- full_join(total_fertility,life_expectancy, 
                 by = c("country","year"))
```

We also want to add the continent for each country. We upload a country to continent map we created.

```{r}
map <- read_delim(file.path("data","continent-info.tsv"),
                  col_names=FALSE,delim ="\t")
continents <- map[[2]][match(dat$country, map[[1]])]
dat <- mutate(dat, continent = continents)
```

We also want to annotate which countries are OECD.
```{r}
oecd_countries <- readRDS(file.path("data","oecd-countries.rds"))
dat <- mutate(dat, oecd = tolower(country) %in% oecd_countries)
```

We will also want to bring in two more pieces of data: the population size:

```{r}
population <- read_csv("http://spreadsheets.google.com/pub?key=phAwcNAVuyj0XOoBL_n5tAQ&output=csv")
names(population)[1] <- "country"
population <- population %>% 
  gather(key = year, value = population,
         2:ncol(population), -country, convert=TRUE)
dat <- full_join(dat, population, by = c("country","year"))
```



Now let's examine these two variables for 1962. The size of the dots will be population size

```{r, warning=FALSE}
library(ggplot2)
theme_set(theme_bw())
p <- ggplot( filter(dat, year==1962) , 
             aes(total_fertility, life_expectancy) )
p + geom_point( aes(color = continent,
                    size=population/10^6)) +
  ggtitle("1962")
```


This data does seems to support a 
western world (Europe) versus the third world (Africa, Asia and South America)  divdie characterized by long life in a small family
and short life in a large family respectively. We can highlight the two groups:

```{r, warning=FALSE}
p + geom_rect(aes(xmin=1.25,xmax=4,ymin=65,ymax=75),
              fill="grey80", alpha=0.8) +
    geom_rect(aes(xmin=5,xmax=8.1,ymin=30,ymax=60),
              fill="grey80", alpha=0.8)+
    geom_point( aes(color = continent,
                    size=population/10^6)) +
  ggtitle("1962")
```


However, this was in 1962. Is it still the case?


```{r, warning=FALSE}
years <- seq(1962,2012,10)
p <- ggplot( filter(dat, year%in%years) , 
             aes(total_fertility, life_expectancy) ) + facet_wrap(~year) + geom_point( aes(color = continent)) 
p
```

Note that by 2012 the dicotomy is no longer as strong. This is particularly true for Asia and Europe. In 1962 they did indeed appear to be divided in two groups, but by 2012 it is no longer the case.

```{r, warning=FALSE}
years <- seq(1962,2012,10)
p <- ggplot( filter(dat, year%in%years & 
                      continent %in% c("Asia","Europe")) , 
             aes(total_fertility, life_expectancy) ) + facet_wrap(~year) + geom_point( aes(color = continent,size=population)) 
p
```


#### Animations
If we want to make an animation we can use this:
```{r, echo=FALSE, warning=FALSE, eval=FALSE}
#library(gganimate)
#p <- ggplot( filter(dat, year>1961 &
##                      !is.na(total_fertility) &
#                    !is.na(life_expectancy) ), 
 #            aes(total_fertility, life_expectancy))+
  #coord_cartesian(ylim = c(30, 85)) +
  #geom_point(aes(color=continent, size=population, 
   #              frame=year))

#gg_animate(p, "output.mp4")
```

## Time Series Plots

We can see the progress across time by creating a time series plot. Here we can see how fertility rages have dropped in Vietnam and now it is similar to the USA.

```{r}
countries <- c("Vietnam","United States")
ggplot( filter(dat,country%in%countries &
                      year > 1959), 
             aes(year,total_fertility,color=country)) +
  geom_line() 
```

The life expectancy has grown in both countries with Vietnam showint a faster rate of improvement.

```{r}
ggplot( filter(dat,country%in%countries &
                      year > 1959), 
             aes(year,life_expectancy,color=country)) +
  geom_line()
```

## Income Distribution

Another common opinion is that wealth distribution across the world has gotten worse and that there is a dichotomy between rich and poor countries. In 1970 this dichotomy can be seen in the distribution of income per capita across countries. Let's create this varialbe and created a density plot that weighs by population size. First download and add GDP to our table.

```{r}
gdp <- read_csv("http://spreadsheets.google.com/pub?key=pyj6tScZqmEfI4sLVvEQtHw&output=csv")
names(gdp)[1] <- "country"
gdp <- gdp %>% 
  gather(key = year, value = gdp,
         2:ncol(gdp), -country, convert=TRUE)
dat <- full_join(dat, gdp, by = c("country","year")) 
 
gdp_per_person <- read_csv("http://spreadsheets.google.com/pub?key=phAwcNAVuyj1jiMAkmq1iMg&output=csv")
names(gdp_per_person)[1] <- "country"
gdp_per_person <- gdp_per_person %>% 
  gather(key = year, value = gdp_per_person,
         2:ncol(gdp_per_person), -country, convert=TRUE)
dat <- full_join(dat, gdp_per_person, by = c("country","year")) 
```

In 1970 the dichotomy is clear
```{r}
ggplot(filter(dat, year==1970), 
       aes(x = gdp_per_person/365)) +
  geom_density(aes(weight=population)) + scale_x_log10()
```

With the majority of world in countires living of $1 a day on average. We can see the 
```{r}
years <- seq(1970,2010,10)
ggplot(filter(dat, year%in%years & !is.na(population)), 
       aes(x = gdp_per_person/365)) +
  geom_density(aes(weight=population)) + scale_x_log10() + facet_grid(~year)
```


```{r}
ggplot(filter(dat, year==2010 & !is.na(population)) %>% mutate(area=ifelse(oecd,"OECD",continent)), 
       aes(x = gdp_per_person/365,fill=area)) +
  geom_density(aes(weight=population),alpha=0.3) +  coord_cartesian(xlim = c(1,150))+scale_x_log10()
```




