---
title: "Insights on Poverty"
date: "January 29, 2016"
output: html_document
---

# Introduction

This lab is based Hans Rosling's talks [New Insights on Poverty](https://www.ted.com/talks/hans_rosling_reveals_new_insights_on_poverty?language=en) and [The Best Stats You've Ever Seen](https://www.ted.com/talks/hans_rosling_shows_the_best_stats_you_ve_ever_seen). 

The lab uses data to answer specific question about global health and economics. The data contradicts commonly held preconceived nations. Let's start by testing your knowledge. For each of the six pairs of countries below, which country do you think  had the highest child mortality in 2015? Which pairs do you think are similar?

1. Sri Lanka or Turkey
2. Poland or South Korea
3. Malaysia or Russia
4. Pakistan or Vietnam
5. Thailand or South Africa

Now let's use data to answer the questions. There are a number of datasets available on [gapminder](http://www.gapminder.org/). You can visit the website and browse the datasets. One of the dataset has to find one containg childhood mortality data.

```{r load libraries,message=FALSE,echo=TRUE}
library(dplyr)
library(readr)
child_mortality <-read_csv("http://spreadsheets.google.com/pub?key=0ArfEDsV3bBwCcGhBd2NOQVZ1eWowNVpSNjl1c3lRSWc&output=csv")
```

We can answer above by simply using the following code:
```{r}
countries <- c("Sri Lanka","Turkey", "Poland", "South Korea", "Malaysia", "Russia","Pakistan","Vietnam","Thailand","South Africa")

answer <- rename(child_mortality, country=`Under five mortality`) %>%
  slice(match(countries,country)) %>% select(country,`2015`) 

bind_cols( slice( answer, seq(1,9,2)), 
           slice( answer, seq(2,10,2)))
```

From here we see that Turkey has a higher rate than Turkey, Poland higher than South Korea, Russia higher than Malaysia, Pakistan much higher than Vietnam and South Africa much higher than Thailand. 

Most people get more than half of these wrong. Why is this? In part it is due to our preconceived notion that the world is divided into two groups: Western world (Turkey, Poland, and Russia are in Europe) versus the third world characterized by long life in a small family
and short life in a large family respectively. Does the data support this dichotomous view of two groups?

The necessary data to answer this question is also available from gapminder. In the following section we will download and organize child mortality, life expectancy, bird rate, populationy, and GDP data for all countries.

# Data Wrangling

We went to the gapminder web page and found the following four relevant

```{r}
urls <- c("http://spreadsheets.google.com/pub?key=0ArfEDsV3bBwCcGhBd2NOQVZ1eWowNVpSNjl1c3lRSWc&output=csv",
          "http://spreadsheets.google.com/pub?key=phAwcNAVuyj2tPLxKvvnNPA&output=csv",
          "http://spreadsheets.google.com/pub?key=phAwcNAVuyj0TAlJeCEzcGQ&output=csv",
          "http://spreadsheets.google.com/pub?key=phAwcNAVuyj0XOoBL_n5tAQ&output=csv",
          "http://spreadsheets.google.com/pub?key=pyj6tScZqmEfI4sLVvEQtHw&output=csv")
names(urls) <- c("child_mortality", "life_expectancy", "fertility","population","gdp")

##Read in all tables into a list
tables <- lapply(urls, read_csv)

##Function to make column name for key uniform and
##change for wide to long format
fix_table <- function(tab, val_name){
  names(tab)[1] <- "country"
  tab <- gather(tab, key=year, value=y, -country, convert = TRUE)
  names(tab)[which(names(tab)=="y")] <- val_name
  return(tab)
}

##Fix each table
for(i in seq_along(tables) ) tables[[i]] <-fix_table(tables[[i]], names(tables)[i])

##Join into one large table
dat <- Reduce(full_join, tables)

rm(tables); gc()
```

We also want to add a column with the continent. In a separate file we have downloaded this information and created the following table:

```{r}
filename <- file.path("data","continent-info.tsv")
map <- read_delim(filename, 
                  col_names = c("country","continent"),
                  delim ="\t")
dat <- left_join(dat, map, continent = continents)
```

We also want to annotate which countries are OECD. We have created a file with this information in separate scripts. We will create a varialbe called group that we will use to determine the type of country: OECD, Asia, Africa or Americas. Note that we leave Ocenia and non-OECD European countries out to avoid having too many groups.

```{r}
oecd <- readRDS(file.path("data","oecd-countries.rds"))
group <- ifelse( dat$country%in%oecd , "OECD", dat$continent  )
group <- ifelse( group%in%c("Europe","Oceania"), NA, group)

dat <- mutate(dat, group=group)
```


# Life expectancy versus birth rate

Let's examine what the world looked like in 1962 in terms of life expectancy versus birth rate. The size of the dots will be population size and the color represents groups of countries.

```{r, warning=FALSE}
library(ggplot2)
theme_set(theme_bw())

filter(dat, year==1962 & !is.na(group) & 
         !is.na(fertility) & !is.na(life_expectancy)) %>%
  ggplot( aes(fertility, y=life_expectancy)) +
  ylab("Life Expectancy") + 
  xlab("Fertility rate (births per woman)") +
  geom_rect(aes(xmin=1.25,xmax=4.25,ymin=65,ymax=85),
              fill="grey80", alpha=0.8) +
  geom_rect(aes(xmin=5,xmax=8.5,ymin=30,ymax=60),
              fill="grey80", alpha=0.8)+
  geom_point( aes(color = group, size=population)) 
```


This data does seems to support a western world (OECD) versus the third world (Africa, Asia and South America)  divdie characterized by long life (y-axis) in a small family (x-axis)and short life in a large family respectively.  

However, this was in 1962. Is it still the case?

```{r, warning=FALSE}
years <- c(1962,2012)
filter(dat, year%in%years & !is.na(group) & 
         !is.na(fertility) & !is.na(life_expectancy)) %>%
  ggplot( aes(fertility, y=life_expectancy)) +
  ylab("Life Expectancy") + 
  xlab("Fertility rate (births per woman)") +
  geom_rect(aes(xmin=1.25,xmax=4.25,ymin=65,ymax=85),
              fill="grey80", alpha=0.8) +
  geom_rect(aes(xmin=5,xmax=8.5,ymin=30,ymax=60),
              fill="grey80", alpha=0.8)+
  geom_point( aes(color = group, size=population)) +
  facet_wrap(~year)
```

Note that by 2012 the dicotomy is no longer as strong. This is particularly true for Asia and Europe. In 1962 they did indeed appear to be divided in two groups, but by 2012 it is no longer the case.

## Animations

We can use the `gganimate` package to create an [animation](animation.html) from 1962 to 2012:
```{r, echo=FALSE, warning=FALSE, eval=FALSE}
library(gganimate)
years <- seq(1962,2012)
p <- filter(dat, year%in%years & !is.na(group) & 
         !is.na(fertility) & !is.na(life_expectancy)) %>%
  ggplot( aes(fertility, y=life_expectancy)) +
  ylab("Life Expectancy") + 
  xlab("Fertility rate (births per woman)") +
  geom_point( aes(color = group, size=population, frame=year))+
  coord_cartesian(ylim = c(30, 85))

gg_animate(p, "output.html")
gg_animate(p, "output.gif", ani.width = 400, ani.height = 350)

```

## Two countries

We can see the progress across time by creating a time series plot. Below we can see how fertility rages have dropped in Vietnam and now it is similar to the USA. Meanwhile, life expectancy has grown in both countries with Vietnam showing a faster rate of improvement.


```{r}
library(cowplot)
countries <- c("Vietnam","United States")
p1 <- filter(dat,country%in%countries & year > 1959) %>% 
  ggplot(aes(year,fertility,color=country)) +
  geom_line() + theme(legend.position="none") +
  annotate("text", x = 1984, y = 5, label = "Vietnam")+
  annotate("text", x = 1970, y = 3, label = "United States")
p2 <- filter(dat,country%in%countries & year > 1959) %>% 
  ggplot(aes(year,life_expectancy,color=country)) +
  geom_line() + theme(legend.position="none")
plot_grid(p1,p2)
```

# Income Distribution

Another common held notionis that wealth distribution across the world has become worse during the last decades and that there is a dichotomy between rich and poor countries. 

In 1970 there was certianly a dichotomy and we can see this by visualizing the income distribution of income per capita across countries. The density plot below is for country per capita incomes but we use weights for larger countries.

We will now add a fifth group, the OPEC countries, which tend to have large average incomes.

```{r}
opec <- scan(file.path("data","opec-countries.txt"),
             what="c", sep="\n")

p <- filter(dat, year==1970 & !is.na(gdp) 
            & !is.na(population) & !is.na(group)) %>%
  mutate(group = ifelse(country%in%opec, "OPEC", group)) %>%
  ggplot(aes(x = gdp/population/365)) +
  xlab("Dollars per day") + 
  theme(axis.line.y=element_blank(),
        axis.text.y=element_blank(),
        axis.title.y=element_blank()) +
  scale_x_log10()
p + geom_density(aes(weight=population/sum(population)))
```

An interpretation of this plot is that there are two clusters of countries: those were the average person earns about $1 dollar a day, and those where they earn about $75. To see what regions these countries are from we can make a density for each.


```{r}
p + geom_density(aes(weight=population/sum(population)),
                 alpha=0.3) +
  aes(fill=group) 
```

```{r}
years <- seq(1970,2010,10)
filter(dat, year%in%years & !is.na(gdp) 
            & !is.na(population) & !is.na(group)) %>%
  mutate(group = ifelse(country%in%opec, "OPEC", group)) %>%
  ggplot(aes(x = gdp/population/365)) +
  xlab("Dollars per day") + 
  theme(axis.line.y=element_blank(),
        axis.text.y=element_blank(),
        axis.title.y=element_blank()) +
  scale_x_log10() +
  geom_density(aes(weight=population/sum(population)),
                 alpha=0.3) + aes(fill=group) +
  facet_grid(.~year)
```

Giving us a new overall density with much less separation between "rich" and "poor":

```{r}
filter(dat, year%in%c(1970,2010) & !is.na(gdp) 
            & !is.na(population) & !is.na(group)) %>%
  mutate(group = ifelse(country%in%opec, "OPEC", group)) %>%
  ggplot(aes(x = gdp/population/365)) +
  xlab("Dollars per day") + 
  theme(axis.line.y=element_blank(),
        axis.text.y=element_blank(),
        axis.title.y=element_blank()) +
  scale_x_log10() + 
  geom_density(aes(weight=population/sum(population))) +
  facet_grid(.~year)
```

# Child Mortality and GDP per capita

We see that the five groups shown here are quite different in income per capita. The plot of life expectancy versus income shows a striking correlation:

```{r}
mutate(dat, group = ifelse(country%in%opec, "OPEC", group),
       survival = 1 - child_mortality/1000) %>%
  filter(year==2010 & !is.na(group)) %>% group_by(group) %>%
  summarise(income=mean(gdp/population,na.rm=TRUE),
            survival=mean(survival,na.rm=TRUE),
            population=sum(population, na.rm=TRUE)) %>%
  ggplot(aes(income,y=survival)) + 
  geom_point(aes(color=group,size=population)) +
  xlab("GDP per capita") +
  ylab("Proportion Child Survival") +
  ggtitle("2010") +
  scale_size_continuous(range = c(5, 18)) + 
  scale_y_continuous(trans=scales::logit_trans(),
                     breaks=c(.95,.97,.99,.995,.997),
                     limits=c(0.79,.998))+
  scale_x_continuous(trans=scales::log10_trans(),
                     limits=c(100,53000),
                     breaks=c(500,5000,50000))
```

However, once we acutally show all the countries in these groups, we see that there is a large amount of variability. We highlight two countries in each group.


```{r}
countries <- c("Sierra Leone", "Mauritius", "Angola","Qatar",
               "Cambodia","Singapore","Chile", "Haiti", 
               "United States","Sweden")
dat2 <- mutate(dat, 
               group = ifelse(country%in%opec, "OPEC", group),
               income = gdp/population,
               survival = 1 - child_mortality/1000) %>%
  mutate(group=ifelse(country=="Chile","Americas",group)) %>%
  filter(year==2010 & 
           !is.na(group) & !is.na(income) & !is.na(survival)) 
ggplot(dat2, aes(income,y=survival)) + 
  geom_point(aes(color=group,size=population)) +
  xlab("GDP per capita") +
  ylab("Proportion Child Survival") +
  ggtitle("2010") +
  scale_y_continuous(trans=scales::logit_trans(),
                     breaks=c(.95,.97,.99,.995,.997),
                     limits=c(0.79,.998))+
  scale_x_continuous(trans=scales::log10_trans(),
                     limits=c(100,53000),
                     breaks=c(500,5000,50000))+
  geom_text(data=filter(dat2, country%in%countries),
            aes(label=country,color=group),  nudge_y = 0.1, 
            show.legend  = FALSE )
```

# Within country variation

Finally, we will show that even within country we have quite a bit of variation. We need to create another dataset which has the within-country income quantiles.

```{r}
#loading income share by population quintiles
urls <- c(
   "http://spreadsheets.google.com/pub?key=pyj6tScZqmEdIyrBS31XAaw&output=csv",
   "http://spreadsheets.google.com/pub?key=tXRyZGCfHsWMmr53VFxrqTw&output=csv",
    "http://spreadsheets.google.com/pub?key=t_-14NtXH6xZX48xHG75z5w&output=csv",
   "http://spreadsheets.google.com/pub?key=twSOUYrIFh2W2snDUt7VaQg&output=csv","http://spreadsheets.google.com/pub?key=tLnCxItXzRSu9gH-5PyEFDw&output=csv")

```

